
---
title: "`HDF5DataFrame`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
package: "`r BiocStyle::pkg_ver('HDF5DataFrame')`"
author: 
  - name: Artur Manukyan
output:
  BiocStyle::html_document
vignette: |
  %\VignetteIndexEntry{HDF5DataFrame}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r load-libs}
library(rhdf5)
library(HDF5Array)
library(HDF5DataFrame)
```

# Introduction

`r Biocpkg("HDF5DataFrame")` is an R/Bioconductor package for HDF5-backed 
DataFrame objects and methods. Each column of a data frame is stored 
as a separate one dimensional array in an HDF5 file. 
`r Biocpkg("HDF5DataFrame")` organizes these arrays and serves them 
as a `DataFrame`-like object.

# Installation

You can install `r Biocpkg("HDF5DataFrame")` from Bioconductor using 
`r CRANpkg("BiocManager")`:

```{r bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("HDF5DataFrame")
```

# Usage

We use `writeHDF5DataFrame` to write the a `data.frame` to an HDF5 file.
You can also specify the HDF5 group (`name`) where columns of the data frame
are stored.

```{r read}
# data
data("chickwts")
metadata <- chickwts

# create HDF5DataFrame and write to hdf5
output_hdf5 <- tempfile(fileext = ".h5")
metadata_hdf5 <- writeHDF5DataFrame(metadata,
                                    filepath = output_hdf5, 
                                    name = "metadata",
                                    replace = TRUE)

# view HDF5DataFrame
metadata_hdf5
```

The `path` method points to the HDF5 file where the data frame is stored. 

```{r path}
path(metadata_hdf5)
```

Each column is stored as a separate one dimensional dataset in the HDF5 file.
Here `metadata` is the group that contains each array with the same length. 

```{r h5ls}
h5ls(output_hdf5)
```

If the HDF5 file already contains a data frame at the specified group with
known name, you can construct an `HDF5DataFrame` object directly

```{r HDF5DataFrame}
metadata_hdf5 <- HDF5DataFrame(filepath = output_hdf5,
                               name = "metadata")
```

# Methods

You can use various `DataFrame` methods on `HDF5DataFrame` objects such as 
`cbind` where an in-memory `data.frame` can be combined with the `HDF5DataFrame`
object. 

```{r methods}
# cbind with in memory data
metadata_hdf5 <- cbind(metadata_hdf5, metadata)
metadata_hdf5
```

You can also realize the `HDF5DataFrame` object into an in-memory 
`data.frame` any time.

```{r methods2}
# coerce to data.frame
metadata_hdf5 <- as.data.frame(metadata_hdf5)
head(metadata_hdf5)
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```